---
title: "Gapminder"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

```{r setup}
library(tidyverse)
library(gapminder)
library(ggstatsplot)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to education people about the state of the world. They promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

__q0__ Perform your "first checks" on the dataset. What variables are in this
dataset?

```{r q0-task}
#?gapminder
#glimpse(gapminder)
summary(gapminder)
```

**Observations**:

Country, continent, year, life expectancy, population and GDP per capita

__q1__ Determine the most and least recent years in the `gapminder` dataset.

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- 2007
year_min <- 1952
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)
print("Nice!")
```

__q2__ Filter on years matching `year_min`, and make a plot of the GDE per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
gapminder %>% 
  filter(year == min(year)) %>%
  group_by(continent) %>% 
  ggplot() + 
  geom_boxplot(aes(x = continent, y = gdpPercap,group = continent)) #+ 
  #ylim(c(0,20000))

```

```{r q2-improvemnet}
gapminder %>% 
  filter(year == min(year)) %>%
  group_by(continent) %>% 
  ggplot() + 
  geom_boxplot(aes(x = continent, y = gdpPercap,group = continent)) + 
  coord_cartesian(
    ylim = c(0,20000)
  )

```

```{r}
gapminder %>% 
  filter(continent == "Oceania") %>% 
  ggplot() + 
  geom_boxplot(aes(x = country, y = gdpPercap, group = country))
```



**Observations**:

There is a single outlier in Asia that completely warps the data and makes most of the information completely off of the scale. The data can be clipped to only show gdp per capita under $20,000, which makes the rest of the data much more visible, but does ignore that one outlier. The GDP per capita in Europe and oceanea are generally higher than the other countries. I'm somewhat surprised that Oceania has the highest GDP per capita average. This may be because of a limited set of countries included in the dataset. This is explained in that the only two countries included in 1952 for Oceania are Australian and New Zealand. 


**Difficulties & Approaches**:

The main challenge was that one country was off of the chart and made the rest of the data unreadable. My approach to solving that was to clip the data to ignore that country, and provide two plots - one that showed all of the data and one that only showed the clipped information. 

__q3__ You should have found at least three outliers in q2. Identify those outliers (figure out which countries they are).

*Note*: I realize the reasons for

```{r q3-task}
gapminder %>% 
  filter(year == min(year)) %>%
  group_by(continent) %>% 
  mutate(quartile1 = quantile(gdpPercap)[2], quartile2 = quantile(gdpPercap)[3] ,quartile3 = quantile(gdpPercap)[4]) %>% 
  mutate(IQR = quartile3 - quartile1) %>% 
  mutate(upperOutlier = (gdpPercap > quartile3 + (IQR * 1.5))) %>% 
  filter(upperOutlier == TRUE) %>% 
  arrange(continent) 
  #mutate(medianGDP = median(gdpPercap)) %>%  
  
    ```
**Observations**:


The outliers are: 
Angola
Gabon
South Africa
United States
Venezuela
Bahrain 
Kuwait 
Switzerland 

It seems that a number of these countries are very oil rich locations with smaller populations. I'm curious to see the median income instead of the average GDP per capita number. This statistic may be highly warped by economic inequalities. It would be interesting to see the median income in the these countries instead of the mean. 




__q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot. Compare the patterns between `year_min` and `year_max`.



*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
outlierCountries = c("Angola","Gabon","South Africa","United States","Venezuela","Bahrain","Kuwait","Switzerland")

gapminderOutlier <- gapminder %>% 
  #filter(year == min(year)) %>%
  group_by(continent,year) %>% 
  mutate(quartile1 = quantile(gdpPercap)[2], quartile2 = quantile(gdpPercap)[3] ,quartile3 = quantile(gdpPercap)[4]) %>% 
  mutate(IQR = quartile3 - quartile1) %>% 
  mutate(upperOutlier = (gdpPercap > quartile3 + (IQR * 1.5))) %>% 
  ungroup() %>%  
  group_by(country) %>% 
  arrange(year, .by_group = TRUE) %>% 
  mutate(earlyOutlier = upperOutlier[1]) %>% 
  select(country, upperOutlier, earlyOutlier, everything())
  




gapminderOutlier %>% 
  filter(year == min(year) | year == max(year)) %>%
  group_by(continent) %>% 
  ggplot() + 
  geom_boxplot(aes(x = continent, y = gdpPercap,group = continent)) + 
  coord_cartesian(
    ylim = c(0,50000)
  ) + 
  geom_point(aes(x = continent, y = gdpPercap, color = earlyOutlier)) +
  facet_grid(.~year)

```


**Observations**:

-Wealth variations between the countries are largely preserved
-Asia and Specifically gain a lot of wealth in the IQR, while Africa and the Americas gain less. 
- Countries that were outliers do not necessarily remain outliers, but generally remain on the upper half of the income range. 



# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

__q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

Question 1: What happens to the outliers GDP per capita over time? Are there large flucations? 

```{r q5-task1}

gapminderOutlier %>% 
  filter(earlyOutlier == TRUE) %>% 
  ggplot() + 
  geom_line(aes(year, gdpPercap, color = country))


```

Obeservations: 
Most of the outlier countries simply consistently become richer over time (as one would expect from inflation). 
Kuwait is unusual in that in drops significantly in the 1970s. Looking into Kuwaiti history, this happened because of a crash in the Kuwaiti stock market and declines in oil prices.  


Question: How is GDP related to life expectancy? 
```{r q5-task2}

gapminder %>% 
  filter(year == min(year)) %>% 
  ggplot() + 
  geom_point(aes(gdpPercap,lifeExp, color = continent)) + 
  geom_smooth(aes(gdpPercap,lifeExp)) +
  coord_cartesian(
    xlim = c(0,20000),
    ylim = c(25,80)
  )
  



```
Note: this plot excludes Kuwait because of the cluster affect it has to the rest of the data. 

Life expectancy does go up with GDP per capita, but it is not a linear function. It is natrual log increase instead - the gains decrease after a country achieves a certain GDP per capita. 

Question 3: Do outlier countries have unusually low populations? 

```{r q5-task3}
gapminderOutlier %>% 
  filter(upperOutlier == TRUE) %>% 
  ggplot() + 
  geom_point(aes(x = gdpPercap,y = pop, color = country)) 

gapminderOutlier %>% 
  filter(upperOutlier == TRUE) %>% 
  filter(country == "Venezuela") %>% 
  ggplot() + 
  geom_point(aes(x = gdpPercap,y = pop, color = country)) + 
  coord_cartesian(
    xlim = c(0,50000),
    ylim = c(0,5e+07)
  )

```
Outlier countries appear to have a mix of large and small populations. However, it is interesting to see how growth in population and wealth over time is related. There appear to be three patterns: 
1) Wealth and population grow at the same rate, leaving GDP per capita unchanged. South Africa is one example of this. 
2) Increases in gdp per capita and population at the same time, indicating that the productivity of the country is increasing. This is likely countries that are industrialized and industrializing.   
3) Increase (or decrease in the case of Kuwait) in GDP per capita without an incrase in population. I would guess that these countries are affected significantly by natural resource prices, especially oil. 
