---
title: "Data Basics"
author: Zach del Rosario
date: 2020-05-03
output: github_document
time: 5
reading: 0
---

*Purpose*: When first studying a new dataset, there are very simple checks we
should perform first. These are those checks.

*Reading*: (None)

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

__q0__ Run the following chunk:

*Hint*: You can do this either by clicking the green arrow at the top-right of
the chunk, or by using the keybaord shortcut `Shift` + `Cmd/Ctrl` + `Enter`.

```{r }
head(iris)
```

This is a *dataset*; the fundamental object we'll study throughout this course.
Some nomenclature:

- The `1, 2, 3, ...` on the left enumerate the **rows** of the dataset
- The names `Sepal.Length`, `Sepal.Width`, `...` name the **columns** of the dataset
- The column `Sepal.Length` takes **numeric** values
- The column `Species` takes **string** values

__q1__ Load the `tidyverse` and inspect the `diamonds` dataset. What do the
`cut`, `color`, and `clarity` variables mean?

*Hint*: You can run `?diamonds` to get information on a built-in dataset.

```{r load-packages}
library(tidyverse)
?diamonds
```
cut: quality of the cut (Fair, Good, Very Good, Premium, Ideal)
color: diamond colour, from D (best) to J (worst)
clarity: a measurement of how clear the diamond is (I1 (worst), SI2, SI1, VS2, VS1, VVS2, VVS1, IF (best))



__q2__ Run `glimpse(diamonds)`; what variables does `diamonds` have?

```{r}
glimpse(diamonds)
```


__q3__ Run `summary(diamonds)`; what are the common values for each of the
variables? How widely do each of the variables vary?

```{r}
summary(diamonds)
```


*Hint*: The `Median` and `Mean` are common values, while `Min` and `Max` give us
a sense of variation.

```{r summary}
summary(diamonds)
```

**Observations**:
carat - seems to be concentrated highly towards the lower end, with a larger number of small diamonds,  with some outliers at the high end
most diamonds are cut well 
Color appears to be realatively normally distributed around the middle classification
Price has some similar characteristics to carat, although it apperas to be less weighted on the lower end. 

- (Write your observations here!)

You should always analyze your dataset in the simplest way possible, build
hypotheses, and devise more specific analyses to probe those hypotheses. The
`glimpse()` and `summary()` functions are two of the simplest tools we have.

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data00-basics-assignment.Rmd)n.
