---
title: "Packages"
# yaml-author
date: 2020-04-30
output: github_document
time: 5
reading: 0
---

*Purpose*: Every time you start an analysis, you will need to *load packages*.
This is a quick warmup to get you in this habit.

*Reading*: (None)

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

__q1__ Create a new code chunk and prepare to load the `tidyverse`.

In [RStudio](https://bookdown.org/yihui/rmarkdown/r-code.html) use the shortcut
`Ctrl + Alt + I` (`Cmd + Option + I` on Mac). Type the command
`library(tidyverse`).

```{r}
library(tidyverse)
```



Make sure to load packages at the *beginning* of your notebooks! This is a
best-practice.

__q2__ Run the chunk with `library(tidyverse)`. What packages are loaded?

In
[RStudio](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts)
press `Ctrl + Alt + T` (`Cmd + Option + T` on Mac) to run the code chunk at your
cursor.


__q3__ What are the main differences between `install.packages()` and `library()`? How often do you need to call each one?

install.packages() adds the package to the system. library() initializes the package in the code so it can be called from that document. install.packages() only needs to be done once on install and library() needs to be called in any document that the library is used in. 


<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-setup01-packages-assignment.Rmd).
